Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-03-20 paper2.qmd
section:all; process:write,figures,code

---
title: "Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation"
author:
  - name: 
      given: Jose Manuel
      family: Rivera Espejo
    orcid: 0000-0002-3088-2783
    url: https://www.uantwerpen.be/en/staff/jose-manuel-rivera-espejo_23166/
    email: JoseManuel.RiveraEspejo@uantwerpen.be
    corresponding: true
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Tine
      family: van Daal
      non-dropping-particle: van
    orcid: https://orcid.org/0000-0001-9398-9775
    url: https://www.uantwerpen.be/en/staff/tine-vandaal/
    email: tine.vandaal@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Sven
      family: De Maeyer
      non-dropping-particle: De
    orcid: 0000-0003-2888-1631
    url: https://www.uantwerpen.be/en/staff/sven-demaeyer/
    email: sven.demaeyer@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Steven
      family: Gillis
    orcid: 
    url: https://www.uantwerpen.be/nl/personeel/steven-gillis/
    email: steven.gillis@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Linguistics
        group: Centre for computational linguistics, psycholinguistics, and sociolinguistics (CLiPS)
funding: 
  statement: "The project was founded through the Research Fund of the University of Antwerp (BOF)."
keywords:
  - causal inference
  - directed acyclic graphs
  - structural causal models
  - bayesian statistical methods
  - thurstonian model
  - comparative judgement
  - probability
  - statistical modeling
abstract: |
  This study revisits Thurstone's law of comparative judgments (CJ) by addressing two key limitations in traditional approaches. Firstly, it addresses the overreliance on the assumptions of Thurstone's Case V in the statistical analysis of CJ data. Secondly, it addresses the apparent disconnect between CJ's approach to trait measurement and hypothesis testing. We put forward a systematic approach based on causal analysis and Bayesian statistical methods, which results in a model that facilitates a more comprehensive understanding of the factors influencing CJ experiments while offering a robust statistical translation. The new model accommodates unequal dispersions and correlations between stimuli, enhancing the reliability and validity of CJ's trait estimation, thereby ensuring the accurate measurement and interpretation of comparative data. The paper highlights the relevance of this updated framework for modern empirical research, particularly in education and social sciences. This contribution advances current research methodologies, providing a robust foundation for future applications in diverse fields.
key-points:
  - (to do)
date: last-modified
bibliography: references.bib
---

@@

<!-- 2. What is the effectiveness of CJ -->
Numerous studies have documented the effectiveness of CJ in assessing traits and competencies over the past decade. These studies have highlighted three aspects of the method's effectiveness: its reliability, validity, and practical applicability. Research on reliability suggests that CJ requires a relatively modest number of pairwise comparisons [@Verhavert_et_al_2019; @Crompvoets_et_al_2022] to generate trait scores that are as precise and consistent as those generated by other assessment methods [@Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023]. In addition, the evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021]. Meanwhile, research on the validity of CJ scores indicates their capacity to represent accurately the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Bartholomew_et_al_2018; @Bouwer_et_al_2023]. Moreover, research on its practical applicability highlights CJ's versatility across both educational and non-educational contexts [@Kimbell_2012; @Jones_et_al_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].

@@

<!-- 3. But what critical issues remain in CJ Research? -->
Nevertheless, despite the increasing number of CJ studies, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved. This study identifies and discusses two prominent issues that can undermine the reliability and validity of CJ's trait estimates. The first issue arises from the excessive reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data. The second stems from the apparent disconnect between CJ's approach to trait measurement and hypothesis testing. This study addresses these issues by extending Thurstone's general form through a systematic and integrated approach based on Causal and Bayesian inference methods.

<!-- 4. what is the organization of the study? -->
As a result, the study divides its content into six main sections. @sec-thurstone_theory provides an overview of Thurstone's theory. @sec-theory-issues discusses the identified issues in detail. @sec-theoretical extends Thurstone's general form to address these challenges. The extension integrates core theoretical principles alongside key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. @sec-statistical translates these theoretical and practical elements into a probabilistic statistical model for the analysis of dichotomous pairwise comparison data. @sec-discussion discusses the findings, limitations, challenges and explores avenues for future research. Finally, @sec-conclusion summarizes the study's key insights.

@@

<!-- 3. Thurstone's Theory of CJ: discriminal difference -->
However, since the individual discriminal processes of the stimuli are not directly observable, the theory introduces the *law of comparative judgment*. This law posits that in pairwise comparisons, a judge perceives the stimulus with a discriminal process positioned further along the trait continuum as possessing more of the trait [@Bramley_2008, pp. 251]. This suggests that the relative distance between stimuli, rather than their absolute positions on the continuum, likely defines the outcome of pairwise comparisons. Indeed, the theory assumes that the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*, determines the observed dichotomous outcome. Furthermore, the theory assumes that because the individual discriminal processes form a Normal distribution on the continuum, the discriminal difference will also conform to a Normal distribution [@Andrich_1978, pp. 452]. In this distribution, the mode (mean) represents the relative separation between the stimuli, and its dispersion indicates the variability of that separation.

@@

<!-- 4. Example of discriminal difference -->
@fig-discriminal_difference illustrates the distribution of the discriminal difference for the hypothetical texts depicted in @fig-discriminal_process. The figure indicates that the judge perceives Text B as having significantly higher quality than Text A. This conclusion is supported by two key observations: the positive difference between their modal discriminal processes $(T_{B} - T_{A} > 0)$ and the probability area where the discriminal difference distinctly favors Text B over Text A, represented by the shaded gray area denoted as $P(B > A)$. As a result, the dichotomous outcome of this comparison is more likely to favor Text B over Text A.

# Two prominent issues in CJ practice {#sec-theory-issues}

<!-- 1. What is the problem with the general form -->
Thurstone noted from the outset that his general form, described in @sec-thurstone_theory, led to a *trait scaling problem*. Specifically, the model required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b, pp. 267]. For instance, in a CJ experiment with five texts, the general form would require estimating $20$ parameters: five modal discriminal processes, five discriminal dispersions, and $10$ correlations --one per comparison (see @tbl-thurstone_cases). However, a single judge could provide only ${5 \choose 2} = 10$ unique comparisons, an insufficient data set to estimate the required parameters.

@@

<!-- 3. But are our specific problems? -->
It is evident that although Thurstone developed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269] and cautioned that its use "should not be made without experimental test" [@Thurstone_1927b, pp. 270], the CJ literature has nevertheless relied the most on it in practice. This preference is particularly notable given that Case V incorporates the most extensive set of simplifying assumptions among all the cases [@Bramley_2008, pp. 253; @Kelly_et_al_2022, pp. 677] (see @tbl-thurstone_cases).

Furthermore, since Thurstone's primary goal was to define a psychological scale and "allocate the compared stimuli on this continuum" [@Thurstone_1927b, pp. 269], he did not provide guidance on how to use these measurements when, for instance, researchers aim to test hypotheses related to the measured traits. Although the CJ practice has circumvented this issue by using scores' point estimates or their transformations for such analyses, a fundamental question remains: Is this method appropriate for addressing these inquiries?

<!-- 4. Section division -->
Thus, this section discusses these two prominent issues, as they could undermine the reliability and validity of CJ's trait estimates [@Perron_et_al_2015, pp. 2]. Specifically, @sec-theory-issue1 examines the the excessive reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data. Moreover, @sec-theory-issue2 focuses on the apparent disconnect between the approaches to trait measurement and hypothesis testing in CJ.

## The overreliance on Case V in the statistical analysis of CJ data {#sec-theory-issue1}

@@

<!-- 1. The Case V and its simpler statistical representation: the BTL model -->
As previously discussed, despite relying on the most extensive set of simplifying assumptions [@Bramley_2008, pp. 253; @Kelly_et_al_2022, pp. 677], Case V remains the most widely used case in the CJ literature. This popularity stems mainly from its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959]. The BTL model mirrors the assumptions of Case V, with one notable distinction: whereas Case V assumes a Normal distribution for the stimuli discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp. 254] (see @tbl-thurstone_cases). As noted by @vanderLinden_et_al_2017_I [pp. 16], this substitution has little impact on the model's estimation or interpretation because the discriminal process scale is arbitrary up to a non-monotonic transformation. Furthermore, this minimal impact is reinforced by the fact that the Normal and Logistic distributions exhibit analogous statistical properties, differing only by a scaling factor of approximately $1.7$.

@@

<!-- 2. But what are limitations of Case V (and the BTL model)? -->
However, as mentioned earlier, Thurstone originally developed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269], prioritizing statistical simplicity over precision in trait measurement [@Kelly_et_al_2022, pp. 677]. He explicitly warned against its untested application, stating that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270]. Furthermore, he acknowledged that some assumptions could prove problematic when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp. 376]. Consequently, given that modern CJ applications frequently involve such traits and stimuli, two main assumptions of Case V and, by extension, of the BTL model may not consistently hold in theory or practice, namely the assumption of equal dispersion and zero correlation between stimuli.

@@

<!-- 1. But does it hold?  -->
Thurstone argued that the assumption of equal dispersions may not apply when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp. 376], as these traits and stimuli can introduce judgment discrepancies due to their unique characteristics [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. Indeed, the CJ literature may already provide evidence of this violation through "misfit" statistics. For example, labeling texts as *misfits* indicates that comparisons involving these texts result in more judgment discrepancies than those involving other texts [@Pollitt_2004, pp. 11; @Pollitt_2012b, pp. 290]. These discrepancies imply that the discriminal differences for *misfit* texts have broader distributions, further suggesting that their discriminal processes exhibit more dispersion than those of other texts. This idea can be illustrated using @fig-discriminal_process, where text B shows a wider discriminal dispersion than text A. As a result, the texts' discriminal processes overlap more, making it difficult for a judge to determine which text possesses more of the trait.

A similar line of reasoning applies to the concept of "misfit" judges, whose evaluations deviate substantially from the shared consensus due to the unique characteristics of the stimuli or the judges themselves [@Pollitt_2012a, pp. 164-165; @Pollitt_2012b, pp. 289-290; @vanDaal_et_al_2016, pp. 4; @Goossens_et_al_2018, pp. 20]. These "misfit" judges and their associated deviations can give rise to additional statistical and measurement issues, which we discuss in more detail in @sec-theory-issue1b.

<!-- Of these, the most questionable is the one that the discriminal dispersions of all the objects are equal. Thurstone clearly did not expect this to hold for any but the most simple stimuli. The scripts used in cross-moderation exercises are obviously far more complex than any used by Thurstone, so it seems naive to expect this assumption to hold here [@Bramley_2008]. -->

<!-- 2. What does the equal dispersion between stimuli imply? -->
According to the theory, discrepancies in the discriminal dispersions of stimuli shape the distribution of the discriminal difference, exerting a direct influence on the outcome of pairwise comparisons. A thought experiment can illustrate this idea. In this experiment, researchers can observe the discriminal processes for the texts depicted in @fig-discriminal_process. Furthermore, the experiment assumes that the discriminal dispersion for Text A remains constant and that the texts are uncorrelated $(\rho=0)$. @fig-dispersion reveals that an increase in the uncertainty associated with the perception of Text B compared to Text A $(\sigma_{B} - \sigma_{A})$, broadens the distribution of their discriminal difference. This broadening affects the probability area where the discriminal difference distinctly favors Text B over Text A, expressed as $P(B>A)$, ultimately influencing the comparison outcome. Additionally, the figure reveals that when the discriminal dispersions of the texts are equal $(\sigma_{B} - \sigma_{A}=0)$, the discriminal difference is more narrow compared to situations where their dispersions differ. Consequently, the discriminal difference is more likely to favor Text B over Text A (shaded gray area).

@@

<!-- 2. But this is not the scenario we observe -->
In experimental practice, however, the thought experiment occurs in reverse. Researchers first observe the comparison outcome and then use the BTL model to infer the discriminal difference between stimuli and their respective discriminal processes [@Thurstone_1927a, pp. 373]. Consequently, the ability of the outcome to reflect *true* differences between stimuli largely depends on the validity of the model's assumptions [@Kohler_et_al_2019, pp. 150], particularly the assumption of equal dispersions. For example, when researchers correctly assume equal dispersions between texts, the BTL model estimates a discriminal difference distribution that accurately represents the *true* discriminal difference between them. This scenario is illustrated in @fig-dispersion, when the model's discriminal difference distribution aligns with the *true* discriminal difference distribution, represented by the thick continuous line corresponding to $\sigma_{B}-\sigma_{A}=0$. The accuracy of this discriminal difference then ensures reliable estimates for the texts’ discriminal processes.

@@

<!-- 2. So, what happens when the assumption does not hold? -->
However, erroneously assuming equal dispersions between stimuli can give rise to significant statistical and measurement issues. Specifically, the model may overestimate the trait's reliability, that is, the degree to which the outcome accurately reflects the *true* discriminal differences between stimuli. This overestimation, in turn, results in spurious conclusions about these differences [@McElreath_2020, pp. 370] and, by extension, about the underlying discriminal processes of stimuli. @fig-dispersion also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the *true* discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$. Additionally, if researchers recognize that misfit statistics highlight these critical differences in dispersions, the conventional CJ practice of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018] can unintentionally discard valuable information, introducing bias into trait estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. The direction and magnitude of these biases are often unpredictable, as they depend on which stimuli are excluded from the analysis.

@@

<!-- 1. But does it hold?  -->
In contrast, Thurstone assumed that stimuli were uncorrelated because judges' biases, arising from two opposing and equally weighted effects occurring during the pairwise comparisons, canceled each other out [@Thurstone_1927b, pp. 268]. @Andrich_1978 provided a mathematical demonstration of this cancellation using the BTL model under the assumption of discriminal processes with additive biases. However, the assumption of zero correlation is not met in practice in at least two scenarios. First, when intricate aspects of multidimensional, complex traits or heterogeneous stimuli influence judges' perceptions, generating biases that resist cancellation. Second, when an additional hierarchical structure is relevant to the stimuli.

<!-- 2. The first scenario where the assumption does not hold -->
In the first scenario, research on text quality suggests that when judges evaluate multidimensional, complex traits or heterogeneous stimuli, they often rely on various intricate characteristics of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. These additional relevant characteristics are unlikely to be equally weighted or opposing. As a result, they may exert an uneven influence on judges' perceptions, creating biases in their judgments that resist cancellation. For example, this could occur when a judge assessing the argumentative quality of a text places more weight on its grammatical accuracy than other judges, thereby favoring texts with fewer errors but weaker arguments. Furthermore, since the discriminal difference of the stimuli becomes an observable outcome only through the judges' perceptions, these biases could ultimately introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II, pp. 346]. While direct evidence for this particular scenario is lacking, studies such as @Pollitt_et_al_2003 demonstrate the presence of such biases, supporting the notion that the factors influencing pairwise comparisons may not always cancel out.

>>>>

---
title: "Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation"
author:
  - name: 
      given: Jose Manuel
      family: Rivera Espejo
    orcid: 0000-0002-3088-2783
    url: https://www.uantwerpen.be/en/staff/jose-manuel-rivera-espejo_23166/
    email: JoseManuel.RiveraEspejo@uantwerpen.be
    corresponding: true
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Tine
      family: van Daal
      non-dropping-particle: van
    orcid: https://orcid.org/0000-0001-9398-9775
    url: https://www.uantwerpen.be/en/staff/tine-vandaal/
    email: tine.vandaal@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Sven
      family: De Maeyer
      non-dropping-particle: De
    orcid: 0000-0003-2888-1631
    url: https://www.uantwerpen.be/en/staff/sven-demaeyer/
    email: sven.demaeyer@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Steven
      family: Gillis
    orcid: 
    url: https://www.uantwerpen.be/nl/personeel/steven-gillis/
    email: steven.gillis@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Linguistics
        group: Centre for computational linguistics, psycholinguistics, and sociolinguistics (CLiPS)
funding: 
  statement: "The project was founded through the Research Fund of the University of Antwerp (BOF)."
keywords:
  - causal inference
  - directed acyclic graphs
  - structural causal models
  - bayesian statistical methods
  - thurstonian model
  - comparative judgement
  - probability
  - statistical modeling
abstract: |
  This study revisits Thurstone's law of comparative judgment (CJ), focusing on two prominent issues of traditional approaches. First, it critiques the heavy reliance on Thurstone's Case V assumptions and, by extension, the Bradley-Terry-Luce (BTL) model when analyzing CJ data. Specifically, the study raises concerns about the assumptions of equal discriminal dispersions and zero correlation between the stimuli. While these assumptions simplify the trait measurement process, they may fail to capture the complexity of CJ data, potentially leading to unreliable and inaccurate trait estimates. Second, the study highlights the apparent disconnect between CJ's trait measurement and hypothesis testing processes. Although separating these processes simplifies the analysis of CJ data, it may also undermine the reliability of various statistical results derived from the measurement.

  To address these issues, the study extends Thurstone's general form using a systematic and integrated approach based on Causal and Bayesian inference methods. This extension integrates core theoretical principles alongside key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. It then translates these elements into a probabilistic statistical model for analyzing dichotomous CJ data, overcoming the rigid assumptions of Case V and the BTL model.

  Finally, the study emphasizes the relevance of this extension for contemporary empirical CJ research, particularly stressing the need for bespoke CJ models tailored to the specific assumptions of the experiments and data. It also lays the foundation for broader applications, encouraging researchers across the social sciences to adopt more robust and interpretable methodologies.
key-points:
  - (to do)
date: last-modified
bibliography: references.bib
---

@@

<!-- 2. What is the effectiveness of CJ -->
Numerous studies have documented the effectiveness of CJ in assessing traits and competencies over the past decade. These studies have highlighted three aspects of the method's effectiveness: its reliability, validity, and practical applicability. Research on reliability suggests that CJ requires a relatively modest number of pairwise comparisons [@Verhavert_et_al_2019; @Crompvoets_et_al_2022] to generate trait scores that are as precise and consistent as those generated by other assessment methods [@Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023]. In addition, the evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021]. Meanwhile, research on validity indicates the capacity of CJ scores to represent accurately the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Bartholomew_et_al_2018; @Bouwer_et_al_2023]. Lastly, research on its practical applicability highlights CJ's versatility across both educational and non-educational contexts [@Kimbell_2012; @Jones_et_al_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].

@@

<!-- 3. But what critical issues remain in CJ Research? -->
Nevertheless, despite the increasing number of CJ studies, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved. This study identifies and discusses two prominent issues of traditional approaches that can undermine the reliability and validity of CJ's trait estimates  [@Perron_et_al_2015, pp. 2]. First, it critiques the heavy reliance on Thurstone's Case V assumptions [@Thurstone_1927b] and, by extension, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] when analyzing CJ data. Specifically, the study raises concerns about the assumptions of equal discriminal dispersions and zero correlation between the stimuli. While these assumptions simplify statistical modeling, they may fail to capture the complexity of CJ data, potentially leading to unreliable and inaccurate trait estimates. Second, the study highlights the apparent disconnect between CJ's trait measurement and hypothesis testing processes. Although separating these processes simplifies the analysis of CJ data, it may also undermine the reliability of various statistical results derived from the measurement.

<!-- 4. what is the organization of the study? -->
To address these issues, the study extends Thurstone's general form using a systematic and integrated approach based on Causal and Bayesian inference methods. As a result, the study divides its content into six main sections. @sec-thurstone_theory provides an overview of Thurstone's theory. @sec-theory-issues discusses the identified issues in detail. @sec-theoretical extends Thurstone's general form to address these challenges. The extension integrates core theoretical principles alongside key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. @sec-statistical translates these theoretical and practical elements into a probabilistic statistical model to analyze dichotomous pairwise comparison data. @sec-discussion discusses the findings, limitations, and challenges and explores avenues for future research. Finally, @sec-conclusion summarizes the study's conclusions.

@@

<!-- 3. Thurstone's Theory of CJ: discriminal difference -->
However, since the individual discriminal processes of the stimuli are not directly observable, the theory introduces the *law of comparative judgment*. This law posits that in pairwise comparisons, a judge perceives the stimulus with a discriminal process positioned further along the trait continuum as possessing more of the trait [@Bramley_2008, pp. 251]. This suggests that pairwise comparison outcomes depend on the relative distance between stimuli, not their absolute positions on the continuum. Indeed, the theory assumes that the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*, determines the observed dichotomous outcome. Furthermore, the theory assumes that because the individual discriminal processes form a Normal distribution on the continuum, the discriminal difference will also conform to a Normal distribution [@Andrich_1978, pp. 452]. In this distribution, the mode (mean) represents the relative separation between the stimuli, and its dispersion indicates the variability of that separation.

@@

<!-- 4. Example of discriminal difference -->
@fig-discriminal_difference illustrates the distribution of the discriminal difference for the hypothetical texts depicted in @fig-discriminal_process. The figure indicates that the judge perceives Text B as having significantly higher quality than Text A. Two key observations support this conclusion: the positive difference between their modal discriminal processes $(T_{B} - T_{A} > 0)$ and the probability area where the discriminal difference distinctly favors Text B over Text A, represented by the shaded gray area denoted as $P(B > A)$. As a result, the dichotomous outcome of this comparison is more likely to favor Text B over Text A.

# Two Prominent Issues in Traditional CJ Practice {#sec-theory-issues}

<!-- 1. What is the problem with the general form -->
Thurstone noted from the outset that his general form, described in @sec-thurstone_theory, led to a *trait scaling problem*. Specifically, the model required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b, pp. 267]. For instance, in a CJ experiment with five texts, the general form would require estimating $20$ parameters: five modal discriminal processes, five discriminal dispersions, and $10$ correlations --one per comparison (see @tbl-thurstone_cases). However, a single judge could only provide ${5 \choose 2} = 10$ unique comparisons, an insufficient data set to estimate the required parameters.

@@

<!-- 3. But are our specific problems? -->
However, Thurstone designed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269], prioritizing statistical simplicity over precise trait measurement. He cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270] because it imposes the most extensive set of simplifying assumptions among all the proposed cases [@Bramley_2008, pp. 253; @Kelly_et_al_2022, pp. 677] (see @tbl-thurstone_cases). Therefore, it is surprising that, despite these limitations, CJ research has predominantly relied on Case V to measure different traits, raising significant concerns about the reliability and validity of such measurements in contexts where the case's assumptions may not hold [@Kelly_et_al_2022, pp. 677].

Furthermore, since Thurstone's primary goal was to define a psychological scale and "allocate the compared stimuli on this continuum" [@Thurstone_1927b, pp. 269], he did not provide specific guidance on how to use the measurement scores from Case V to conduct hypothesis tests related to the measured traits. Although the CJ practice has circumvented this issue by using the point estimates of the scores or their transformations for such analyses, a fundamental question remains: Is this method appropriate for addressing these inquiries?

<!-- 4. Section division -->
Thus, this section discusses these two prominent issues. Specifically, @sec-theory-issue1 examines the heavy reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data. Conversely, @sec-theory-issue2 focuses on the apparent disconnect between the approaches to trait measurement and hypothesis testing in CJ.

## The Case V and the statistical analysis of CJ data {#sec-theory-issue1}

@@

<!-- 1. The Case V and its simpler statistical representation: the BTL model -->
Despite its limitations, Case V remains the most widely used model in CJ literature. This preference primarily stems from the BTL model, which provides a simplified statistical representation of the case. The BTL model mirrors the assumptions of Case V, with one notable distinction: while Case V assumes a Normal distribution for the stimuli' discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp. 254] (see @tbl-thurstone_cases). However, as @vanderLinden_et_al_2017_I [pp. 16] noted, this substitution has minimal impact on the model's estimation or interpretation because the discriminal process scale is arbitrary up to a non-monotonic transformation. Furthermore, this limited impact is supported by the fact that the Normal and Logistic distributions exhibit analogous statistical properties, differing only by a scaling factor of approximately $1.7$.

@@

<!-- 2. But what are limitations of Case V (and the BTL model)? -->
However, Thurstone acknowledged that some assumptions of Case V could be problematic when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp. 376]. Thus, given that modern CJ applications often involve such traits and stimuli, two key assumptions of Case V, and by extension, the BTL model, may not always hold in theory or practice. These assumptions are the equal dispersion and zero correlation between stimuli.

@@

<!-- 1. What does the equal dispersion between stimuli imply? -->
According to the theory, discrepancies in the discriminal dispersions of stimuli shape the distribution of the discriminal difference, directly influencing the outcome of pairwise comparisons. A thought experiment can help illustrate this idea. In this experiment, researchers observe the discriminal processes for two texts, similar to those shown in @fig-discriminal_process. Additionally, the discriminal dispersion for Text A remains constant, and the texts are uncorrelated  $(\rho=0)$. @fig-dispersion demonstrates that an increase in the uncertainty associated with the perception of Text B relative to Text A $(\sigma_{B} - \sigma_{A})$, broadens the distribution of their discriminal difference. This broadening affects the probability area where the discriminal difference distinctly favors Text B over Text A, expressed as $P(B>A)$, ultimately influencing the comparison outcome. Additionally, the figure reveals that when the discriminal dispersions of the texts are equal, as in the BTL model $(\sigma_{B} - \sigma_{A}=0)$, the discriminal difference is more narrow than when the dispersions differ. As a result, the discriminal difference is more likely to favor Text B over Text A, as it is represented by the shaded gray area.

@@

<!-- 2. But this is not the scenario we observe -->
In experimental practice, however, the thought experiment occurs in reverse. Researchers first observe the comparison outcome and then use the BTL model to infer the discriminal difference between stimuli and their respective discriminal processes [@Thurstone_1927a, pp. 373]. Consequently, the outcome's ability to reflect *true* differences between stimuli largely depends on the validity of the model's assumptions [@Kohler_et_al_2019, pp. 150], in this case, the assumption of equal dispersions. For instance, when the assumption accurately captures the complexity of the data, the BTL model estimates a discriminal difference distribution that accurately represents the *true* discriminal difference between the texts. This scenario is illustrated in @fig-dispersion, when the model's discriminal difference distribution aligns with the *true* discriminal difference distribution, represented by the thick continuous line corresponding to $\sigma_{B}-\sigma_{A}=0$. The accuracy of this discriminal difference then ensures reliable estimates for the texts’ discriminal processes.

@@

<!-- 3. But does it hold in practice?  -->
Notably, while the assumption of equal dispersions simplifies the trait measurement process, evidence from the CJ literature suggests that the BTL model may fail to capture the complexity of modern CJ data. Specifically, the assumption of equal dispersions may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp. 376; @Bramley_2008, pp. 273; @Kelly_et_al_2022, pp. 678], as these traits and stimuli can introduce judgment discrepancies due to their unique characteristics [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. Indeed, the CJ literature may already provide evidence of such discrepancies, particularly in the form of "misfit" statistics. For instance, *misfit texts* are those whose comparisons result in more judgment discrepancies than those involving other texts [@Pollitt_2004, pp. 11; @Pollitt_2012b, pp. 290].

<!-- Of these, the most questionable is the one that the discriminal dispersions of all the objects are equal. Thurstone clearly did not expect this to hold for any but the most simple stimuli. The scripts used in cross-moderation exercises are obviously far more complex than any used by Thurstone, so it seems naive to expect this assumption to hold here [@Bramley_2008, pp. 273]. -->

<!-- It also seems unclear whether Thurstone’s work really can be applied to the stimuli typically used in assessment. While Thurstone used his Law to create measurement scales for purely psychological or subjective entities, he did so only with stimuli whose values could be evaluated relatively instantaneously, such as short statements and handwriting samples. He was also sceptical about its use for very heterogeneous stimuli (Thurstone, 1927b) [@Kelly_et_al_2022, pp. 678]. -->

<!-- 4. How these judgment discrepancies can happen? -->
These discrepancies may arise from two possible causes. First, the discriminal processes of misfit texts may substantially overlap with those of non-misfit texts. Second, misfit texts may represent outlying observations with characteristics that make them difficult to compare and position on the trait scale. Regarding the first cause, the overlap can occur when misfit texts share the same trait space and have equal dispersion as non-misfit texts or when they exhibit larger discriminal dispersions. Although the BTL model can distinguish the first scenario, it is inherently agnostic to the second. @fig-discriminal_process illustrates how an increased discriminal dispersion can widen the overlap between two texts, leading to more judgment discrepancies. In the figure, text B has a broader discriminal dispersion than text A, expanding the overlap between their discriminal processes, where judges struggle to determine which text possesses more of the trait.

<!-- A similar challenge arises in the first scenario, where overlapping trait spaces contribute to judgment inconsistencies. -->

<!-- 5. So, what happens when the assumption does not hold? -->
Thus, erroneously assuming equal dispersions between stimuli can give rise to significant statistical and measurement issues. Specifically, the BTL model may overestimate the trait's reliability, that is, the degree to which the outcome accurately reflects the *true* discriminal differences between stimuli. This overestimation, in turn, results in spurious conclusions about these differences [@McElreath_2020, pp. 370] and, by extension, about the underlying discriminal processes of stimuli. @fig-dispersion also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the *true* discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$.

Finally, regarding the second cause, if researchers recognize that misfit statistics may represent outlying observations, the conventional CJ practice of  excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], may unintentionally discard valuable information, introducing bias into the trait estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. The direction and magnitude of these biases remain unpredictable, as they depend on which stimuli researchers exclude from the analysis.

@@

<!-- 1. What does the zero correlation between stimuli imply? -->
The correlation $\rho$ measures how much the judges' perception of a specific trait in one stimulus depends on their perception of the same trait in another. As with the discriminal dispersions, this correlation shapes the distribution of the discriminal difference, directly impacting the outcomes of pairwise comparisons. A similar thought experiment, as the one depicted in @sec-theory-issue1a, can illustrate this idea. The experiment only assumes that the discriminal dispersions for both texts remain constant. @fig-correlation reveals that as the correlation between the texts increases, the distribution of their discriminal difference becomes narrower. This narrowing affects the area under the curve where the discriminal difference distinctly favors Text B over Text A, denoted as $P(B > A)$, thus influencing the comparison outcome. Furthermore, the figure shows that when two texts are independent or uncorrelated $(\rho=0)$, their discriminal difference is less narrow compared to scenarios where the texts are positively correlated. As a result, the discriminal difference is less likely to favor Text B over Text A, as it is represented by the shaded gray area.

<!-- 2. But this is not the scenario we observe -->
As outlined in the previous section, researchers approach this process in reverse. They begin by observing the outcomes and using the BTL model to estimate the stimuli' discriminal differences and discriminal processes. Given that the BTL model assumes independent discriminal processes across comparisons, if this assumption holds in the data, the model estimates a discriminal difference distribution that accurately reflects the *true* discriminal difference between the texts. Once more, the accuracy of the discriminal difference ensures reliable estimates for the discriminal processes of the texts.
